### 2022-02-27 doc/src/DuoPLIC.tex

The Advanced PLIC of the previous chapter is not backward compatible
with the original {\RISCV} PLIC.%
\footnote{%
A proposed standard for the original PLIC is the document
\textit{{\RISCV} Platform-Level Interrupt Controller Specification},
available as of mid-2021 here:
\z{https://github.com/{\allowbreak}riscv/{\allowbreak}riscv-plic-spec}.
However, both free implementations of the PLIC and those of some major
{\RISCV} vendors such as SiFive have been observed to differ is small
but significant ways from this document and from each other.
Some important details of the original PLIC were better documented in
older versions of
\textit{The {\RISCV} Instruction Set Manual, Volume~II: Privileged
Architecture}, in a chapter titled
``Platform-Level Interrupt Controller (PLIC).''
That chapter has since been removed from more recent versions of
\textit{The {\RISCV} Instruction Set Manual}.%
}
To ease the transition of software from the original PLIC to the
Advanced PLIC, a {\RISCV} system may have a \emph{\mbox{Duo-PLIC\/}}
that software can configure to act as either form of PLIC.
A \mbox{Duo-PLIC} is an extension of an Advanced PLIC, having all the
usual memory-mapped interfaces of an Advanced PLIC, plus, at some other
memory addresses, the memory-mapped interface of an original {\RISCV}
PLIC too.
When the \mbox{Duo-PLIC} has \emph{compatibility mode} enabled,
interrupts from the \mbox{Duo-PLIC} to harts are controlled exclusively
by the original PLIC side of the \mbox{Duo-PLIC};
and when compatibility mode is disabled, interrupts to harts
are controlled exclusively by the Advanced PLIC side of the
\mbox{Duo-PLIC}.
A \mbox{Duo-PLIC} comes out of reset with compatibility mode enabled,
so older software with no knowledge of an Advanced PLIC can see the
\mbox{Duo-PLIC} as an original PLIC.

>>>>

The Advanced PLIC of the previous chapter is not backward compatible
with the original {\RISCV} PLIC.%
\footnote{%
A proposed standard for the original PLIC is the document
\textit{{\RISCV} Platform-Level Interrupt Controller Specification},
available as of mid-2021 here:
\z{https://github.com/{\allowbreak}riscv/{\allowbreak}riscv-plic-spec}.
However, both free implementations of the PLIC and those of some major
{\RISCV} vendors such as SiFive have been observed to differ in small
but significant ways from this document and from each other.
Some important details of the original PLIC were better documented in
older versions of
\textit{The {\RISCV} Instruction Set Manual, Volume~II: Privileged
Architecture}, in a chapter titled
``Platform-Level Interrupt Controller (PLIC).''
That chapter has since been removed from more recent versions of
\textit{The {\RISCV} Instruction Set Manual}.%
}
To ease the transition of software from the original PLIC to the
Advanced PLIC, a {\RISCV} system may have a \emph{\mbox{Duo-PLIC\/}}
that software can configure to act as either form of PLIC.
A \mbox{Duo-PLIC} is an extension of an Advanced PLIC, having all the
usual memory-mapped interfaces of an Advanced PLIC, plus, at some other
memory addresses, the memory-mapped interface of an original {\RISCV}
PLIC too.
When the \mbox{Duo-PLIC} has \emph{compatibility mode} enabled,
interrupts from the \mbox{Duo-PLIC} to harts are controlled exclusively
by the original PLIC side of the \mbox{Duo-PLIC};
and when compatibility mode is disabled, interrupts to harts
are controlled exclusively by the Advanced PLIC side of the
\mbox{Duo-PLIC}.
A \mbox{Duo-PLIC} comes out of reset with compatibility mode enabled,
so older software with no knowledge of an Advanced PLIC can see the
\mbox{Duo-PLIC} as an original PLIC.

### 2022-02-27 doc/src/IOMMU.tex

\item
For each I/O device connected to the system through an I/O MMU,
software can configure at the I/O MMU a \emph{device context}, which
associates with the device a specific virtual address space and any
other per-device parameters the I/O MMU may support.
By giving devices each their own seperate device context at an I/O MMU,
each device can be individually configured for a different software
master, usually a guest OS or the main (host) OS.
On every memory access made from a device, hardware indicates to
the I/O MMU the originating device by some form of unique device
identifier, which the I/O MMU uses to locate the appropriate device
context within data structures supplied by software.
For PCI, for example, the originating device may be identified by the
unique triple of PCI bus number, device number, and function number.

>>>>

\item
For each I/O device connected to the system through an I/O MMU,
software can configure at the I/O MMU a \emph{device context}, which
associates with the device a specific virtual address space and any
other per-device parameters the I/O MMU may support.
By giving devices each their own separate device context at an I/O MMU,
each device can be individually configured for a different software
master, usually a guest OS or the main (host) OS.
On every memory access made from a device, hardware indicates to
the I/O MMU the originating device by some form of unique device
identifier, which the I/O MMU uses to locate the appropriate device
context within data structures supplied by software.
For PCI, for example, the originating device may be identified by the
unique triple of PCI bus number, device number, and function number.

### 2022-02-27 doc/src/IPIs.tex

The same kind of mechanism (but with a different set of addresses) may
or may not exist for signalling supervisor-level software interrupts
(major code~1) at remote harts as well.
If not directly supported in this way, then a supervisor-level software
interrupt is typically sent to another hart through an environment call
from supervisor mode to machine mode.
An operating system running in \mbox{S-mode} thus invokes a specific
SBI function for delivering a software interrupt to another hart,
causing machine-level software at the originating hart to send a
machine-level IPI to the destination hart, where software then sets the
supervisor-level software interrupt-pending bit (SSIP) in CSR \z{mip}.

When harts have IMSICs, instead of using the Privileged Architecture's
mechanism for signalling software interrupts at remote harts, an IPI
can be sent to a hart by writing to the destination hart's IMSIC, the
same as a regular message-signalled interrupt (MSI)\@.
In that case, an incoming IPI appears at the destination hart as an
\emph{external interrupt} routed through the IMSIC, rather than as a
software interrupt as before.
However, so long as the same software (e.g.\@ an operating system or
machine monitor) is in control at both endpoints of an IPI, source
and destination, there should be no reason for a destination hart
to misinterpret the purpose of an incoming external interrupt that
represents an IPI.

If harts do not have IMSICs, then the method specified by the {\RISCV}
Privileged Architecture is assumed to be used for IPIs, with software
interrupts signalled at destination harts.
On the other hand, when harts have IMSICs, the machinery for triggering
software interrupts at remote harts is redundant with the capabilities
of the IMSICs, so it is downgraded from a requirement to an option,
useful perhaps only to maximize software compatibility across a range
of {\RISCV} systems, with and without IMSICs.
If a machine implements IMSICs and not the earlier software-interrupt
mechanism, then the bits of CSRs \z{mip} and \z{mie} for machine-level
software interrupts, MSIP and MSIE, are hardwired to zero in harts.

>>>>

The same kind of mechanism (but with a different set of addresses) may
or may not exist for signaling supervisor-level software interrupts
(major code~1) at remote harts as well.
If not directly supported in this way, then a supervisor-level software
interrupt is typically sent to another hart through an environment call
from supervisor mode to machine mode.
An operating system running in \mbox{S-mode} thus invokes a specific
SBI function for delivering a software interrupt to another hart,
causing machine-level software at the originating hart to send a
machine-level IPI to the destination hart, where software then sets the
supervisor-level software interrupt-pending bit (SSIP) in CSR \z{mip}.

When harts have IMSICs, instead of using the Privileged Architecture's
mechanism for signaling software interrupts at remote harts, an IPI
can be sent to a hart by writing to the destination hart's IMSIC, the
same as a regular message-signaled interrupt (MSI)\@.
In that case, an incoming IPI appears at the destination hart as an
\emph{external interrupt} routed through the IMSIC, rather than as a
software interrupt as before.
However, so long as the same software (e.g.\@ an operating system or
machine monitor) is in control at both endpoints of an IPI, source
and destination, there should be no reason for a destination hart
to misinterpret the purpose of an incoming external interrupt that
represents an IPI.

If harts do not have IMSICs, then the method specified by the {\RISCV}
Privileged Architecture is assumed to be used for IPIs, with software
interrupts signaled at destination harts.
On the other hand, when harts have IMSICs, the machinery for triggering
software interrupts at remote harts is redundant with the capabilities
of the IMSICs, so it is downgraded from a requirement to an option,
useful perhaps only to maximize software compatibility across a range
of {\RISCV} systems, with and without IMSICs.
If a machine implements IMSICs and not the earlier software-interrupt
mechanism, then the bits of CSRs \z{mip} and \z{mie} for machine-level
software interrupts, MSIP and MSIE, are hardwired to zero in harts.
